Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study

### 2021-10-14 submission/manuscript.rmd
rmemove all text and restart methods

\newpage
# Methods
## Study design
We ran our analyses on three US observational databases mapped to OMOP-CDM
version. More specifically:

@@

Our cohorts included female new users of teriparatide or oral bisphosphonates
above the age of 65 [@Ryan2013]. Patients were considered to be new-users if
they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab,
abaloparatide, romosozumab) 365 days prior to treatment initiation with
teriparatide or an oral bisphosphonate. We required patients to have at least
one year of observation period before first exposure and an established
osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist,
spine or shoulder/humerus in their recorded medical history.

@@

```{r, echo=FALSE}
incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds")) %>%
  dplyr::filter(analysisType == "matchOnPs_1_to_4") %>%
  dplyr::group_by(database) %>%
  dplyr::summarise(
    maxTreatment = max(treatmentPersons),
    maxComparator = max(comparatorPersons),
    .groups = "drop"
  )
```

<!-- Need to fix the numbers. See icpe abstract -->
A total of `r sum(incidenceOverall$maxTreatment) +
sum(incidenceOverall$maxComparator)` patients were initially considered for our
analyses. Of those, `r sum(incidenceOverall$maxTreatment)` received
teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral
bisphosphonate (Table\@ref(tab:table1)).

```{r, echo=FALSE}
metaCalibrateOverall <- readRDS(
  here::here("data/processed/metaCalibrateOverall.rds")
) %>%
  mutate(
    hr = d(hr),
    lower = d(lower),
    upper = d(upper)
  )
```

The random effects meta-analysis of teriparatide compared to oral
bisphosphonates across the 3 considered databases gave a calibrated overall
hazard ratio of `r metaCalibrateOverall %>% filter(outcome == 101) %>%
pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(lower)`
to `r metaCalibrateOverall %>% filter(outcome == 101) %>% pull(upper)`; 95% CI)
for hip fracture, `r metaCalibrateOverall %>% filter(outcome == 102) %>%
pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(lower)`
to `r metaCalibrateOverall %>% filter(outcome == 102) %>% pull(upper)`; 95%CI)
for major osteoporotic fracture and `r metaCalibrateOverall %>% filter(outcome
== 103) %>% pull(hr)` (`r metaCalibrateOverall %>% filter(outcome == 103) %>%
pull(lower)` to `r metaCalibrateOverall %>% filter(outcome == 103) %>%
pull(upper)`; 95%CI) for vertebral fracture (Figure \@ref(fig:plotMeta)).

```{r, echo=FALSE} 
metaCalibrateRiskStratified <- readRDS(
  here::here("data/processed/metaCalibrateRiskStratified.rds")
) %>%
  mutate(
    hr = d(hr),
    lower = d(lower),
    upper = d(upper)
  )

calibrateRiskStratified <- readRDS(
  here::here("data/processed/calibrateRiskStratified.rds")
) %>%
  mutate(
    hr = d(exp(logRr)),
    lower = d(exp(logLb95Rr)),
    upper = d(exp(logUb95Rr))
  )

hipFractureAbsolute <- readRDS(
  here::here("data/processed/hipFractureAbsolute.rds")
) %>%
  filter(riskStratum == "Q4") %>%
  mutate(
    estimate = d(100 * estimate),
    lower    = d(100 * lower),
    upper    = d(100 * upper)
  )

hipRiskPerformance <- readRDS(here::here("data/raw/predictionPerformance.rds")) %>%
  filter(
    cohort == "EntirePopulation",
    stratOutcome == 101
  )

```

The internally developed prediction models for hip fracture showed adequate performance achieving AUC of `r hipRiskPerformance %>% filter(database == "mdcr") %>% pull(auc) %>% d()`, `r hipRiskPerformance %>% filter(database == "optum_dod") %>% pull(auc) %>% d()` and `r hipRiskPerformance %>% filter(database == "panther") %>% pull(auc) %>% d()` in MDCR, Optum-DOD and Optum-EHR respectively. When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, Optum-DOD and Optum-EHR respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases. For the secondary outcomes (major osteoporotic fracture and vertebral fracture) we found no evidence of risk-based treatment effect heterogeneity. This was anticipated as overall effects of teriparatide on theses outcomes were not significant at the 95% confidence level. The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.

<!-- How much sense does it make to look at meta-analytic hazard ratios in risk-quarters? Different models were used in each case  -->

<!-- For the assessment of treatment effect heterogeneity none of the internally developed prediction models achieved high discrimination, with the models predicting the composite major osteoporotic fracture suffering the most (##Table 2). However, models on hip fracture--the primary efficacy outcome--displayed adequate performance. More information on the performance of the prediction models can be found in the supplement. -->

<!-- When analyzing the absolute risk difference of teriparatide compared to oral bisphosphonates within strata of predicted hip fracture risk we did not observe any clear pattern except for a non-significant absolute benefit for hip fracture within the highest risk patients (Figure 4), namely 0.55% (-0.89 to 1.98; 95%CI) in MDCR, 0.97% (-0.27 to 2.22; 95%CI) in OPTUM_DOD and 0.45% (-0.56 to 1.46; 95%CI) in PANTHER. Absolute and relative treatment effect estimates within strata of predicted risk of the secondary outcomes are available in the supplement.  -->

@@

\newpage
# Tables and figures
```{r table1, echo=FALSE, results='asis'}
incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
incidence <- incidenceOverall %>% 
  filter(analysisType == "matchOnPs_1_to_4") %>%
  mutate(
    database = case_when(
      database == "mdcr" ~ "MDCR",
      database == "optum_dod" ~ "Optum-DOD",
      database == "panther" ~ "Optum-EHR"
    ),
    treatmentDays = round(treatmentDays / 365),
    comparatorDays = round(comparatorDays / 365)
  ) %>%
  select(
    database,
    outcomeId,
    treatmentPersons,
    treatmentDays,
    treatmentOutcomes,
    comparatorPersons,
    comparatorDays,
    comparatorOutcomes
  ) %>%
  arrange(database, outcomeId) %>%
  mutate(
    database = ifelse(
      outcomeId == 101,
      database,
      ""
    ),
    outcomeId = case_when(
      outcomeId == 101 ~ "Hip fracture",
      outcomeId == 102 ~ "Major osteoporotic\nfracture",
      outcomeId == 103 ~ "Vertebral fracture"
    ),
  )

knitr::kable(
  incidence,
  format   = "latex",
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Database",
    "Outcome",
    "Persons",
    "Person years",
    "Events",
    "Persons",
    "Person years",
    "Events"
  ),
  format.args = list(
    big.mar = ","
  ),
   caption = "Number of patients, person years and outcomes for each outcome across all databases."
) %>%
  add_header_above(c("", "", "Teriparatide" = 3, "Bisphosphonates" = 3))
  # kableExtra::add_indent(c(1, 4, 7))
```

\newpage
```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls. Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
knitr::include_graphics(here::here("figures/plotMeta.png"))
```

\newpage
```{r plotAbsoluteHip, echo=FALSE, fig.cap="Absolute risk differences wihtin quarters of predicted hip fracture risk across the 3 considered databases.", out.width = '100%'}
knitr::include_graphics(here::here("figures/plotAbsoluteHip.png"))
```

>>>>

\newpage
# Methods
## Study design

We performed a retrospective new user cohort study to estimate the effectiveness
of teriparatide compared to oral bisphosphonates in patients with
osteoporosis [@Ryan2013]. We included female participants above the age of 50
with established osteoporosis (any condition occurrence of hip, wrist, spine or
shoulder/humerus fracture in their history, prior to treatment initiation) and
at least 365 days of continuous observation period before the index event.
Patients were considered to be new-users if they received no anti-osteoporosis
drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days
prior to treatment initiation with teriparatide or an oral bisphosphonate.

As a secondary analysis, we evaluated treatment effect heterogeneity using a risk-based approach. The

## Data sources

We ran our analyses on three US observational databases mapped to OMOP-CDM
version. More specifically:

@@

@@

@@

\newpage
# Tables and figures
